Signatures of Foreground Sky in Power Spectra of Redshifted Neutral Hydrogen from the Epoch of Reionization

### 2014-09-05 foreground_delay_spectrum_modeling.tex

We characterize signatures of foreground radio emission and instrument configuration on the observed power spectra of redshifted H{\sc i} 21~cm line emission from the epoch of reionization (EoR) using an antenna pair based ``delay spectrum'' analysis technique. We use the 128-tile Murchison Widefield Array (MWA) instrument configuration for our studies. % Our analysis is centered around 185~MHz with 30.72~MHz bandwidth at 80~kHz frequency resolution. 
We simulate the delay spectrum response of interferometers, for the first time, to an all-sky foreground model consisting of diffuse and compact Galactic and extragalactic emission. By tying model parameters to those of a subset of first season of observing with the MWA, our simulations agree with the observed data to within expected uncertainty. A wedge--shaped region of foreground contamination and a spillover from this contamination into relatively foreground--free regions ({\it EoR window}) in the delay spectrum due to spectral properties of foreground emission and the instrument is confirmed. Due to shortening of antenna spacings caused by projection effects, diffuse emission produces edge--heavy {\it two--pronged fork}--like features even on wide antenna spacings, hitherto not predicted. Features in the inner regions of the wedge are predominantly due to compact foreground objects. This results in a {\it pitchfork}--shaped delay spectrum. The bright Galactic center severly contaminates the {\it EoR window} on certain antenna spacings in certain observations. For such cases, we present an effective practical tool for screening interferometers with high foreground contamination and mitigating their effects on EoR data analysis.

@@

One of the key challenges in the statistical detection of power spectrum of the redshifted H{\sc i} signal arises from the contamination by Galactic and extragalactic foregrounds \citep[see, e.g.,][]{dim02,zal04,fur06}. \citet{mor04} show that the inherent isotropy and symmetry of the EoR signal in frequency and spatial wavenumber ($k$) space make it distinguishable from sources of contamination which lack such symmetry. Foreground removal techniques rely on the spectral smoothness of foreground emission \citep{mor06,bow09,liu11,par12,dil13}.

Precise understanding of these foreground signatures is necessary in order to remove foreground systematics from the measured power spectrum. Considerable effort has been made in mapping the residual foreground signatures in observed power spectrum \citep{thy13,pob13,mor12,tro12,dat10,bow09} after foreground subtraction.

In this paper, in one of the most detailed characterizations to date, using an interferometer (antenna pair) based delay spectrum technique, we demonstrate the signatures of unsubtracted diffuse and compact foreground models in delay spectrum, a quantity closely related to the sought power spectrum. The simplicity of this technique and usage of basic interferometric visibility data fourier transformed along frequency (delay spectrum) on individual interferometers offer very effective diagnostics for scheduling observations and data analysis, and an insightful view onto foreground signatures in the EoR H{\sc i} power spectrum. We explore the diversity in these signatures as a function of observation and instrument parameters and confirm our findings with data obtained from the MWA. This simple technique allows us to predict foreground signatures, hitherto not predicted.

We have organized this paper as follows. \S\ref{sec:delay-spectrum} defines the delay spectrum and its relation to the power spectrum. \S\ref{sec:instrument} and \S\ref{sec:obsparms} describe the MWA instrumental parameters and observation parameters used in this study respectively. Analysis of data obtained with the MWA is described in \S\ref{sec:data-analysis}. In \S\ref{sec:modeling}, we describe the building blocks of our delay spectrum modeling including the models used for the antenna power pattern and foreground. Here, we also establish that the results of modeling and data analysis agree with each other. In \S\ref{sec:delay-spectrum-analysis}, we present a detailed analysis of the signatures of foregrounds on the modeled delay spectrum and its dependence on instrumental and observation parameters. In \S\ref{sec:fg-grading}, we use our understanding to grade interferometers based on the amount of foreground contamination estimated by the delay spectrum technique. In \S\ref{sec:summary}, we present our summary and conclusions.

@@

In summary, $V_\eta(\overline{\mathbf{u}},\eta)$, is closely related to visibilities, $V_f(\overline{\mathbf{u}},f)$, the basic data blocks measured by an interferometer, as seen from equation~\ref{eqn:delay-transform}. It captures all the effects of EoR H{\sc i} signal corruption caused by foregrounds and the instrument. At the same time, it is very closely related to the sought power $|V(\overline{\mathbf{k}}_\perp,k_\parallel)|^2$ in $k$--modes containing critical information about spatial scales. For our analysis, we use an equivalent quantity, $V_\eta(\overline{\mathbf{x}},\eta) \equiv V_\eta(\lambda\overline{\mathbf{u}},\eta)$. For purposes of illustration, we often use a simplified quantity $|V_\eta(|\overline{\mathbf{x}}|,\eta)|$.

@@

\section{Instrument Parameters}\label{sec:instrument}

We use the 128-tile layout of the MWA \citep{bea12} and choose a center frequency of 185~MHz with a bandwidth of 30.72~MHz divided into 384 frequency channels each of width 80~kHz. The choice of center frequency translates to $z\approx 6.68$ for the 21~cm spin flip transition of H{\sc i}. The bandwidth was chosen to be close to the maximum instantaneous value of the MWA in order to maximize the resolution in delay coordinates and obtain detailed characteristics of delay spectra from foregrounds. We restrict our analysis to the innermost antenna spacings with $|\overline{\mathbf{x}}| \leq x_\textrm{max} = 200$~m, or $|\overline{\mathbf{u}}| \leq u_\textrm{max} = 123$ at a frequency of 185~MHz. These interferometers are responsive to transverse spatial structures on scales $\gtrsim 1/u_\textrm{max} \approx 28$\arcmin. Hereafter, we use the terms tile and antenna interchangeably to refer to the MWA tile.

The bandpass weights in the frequency window of the MWA are obtained through a 8-tap Polyphase Filter Bank (PFB) using a Kaiser window with the $\beta$ parameter set to 5. The bandpass weights consist of coarse frequency channels of width 1.28~MHz each of which contains 16 finer frequency channels of width 80~kHz. Our analysis, therefore, contains 24 coarse channels. As the data are subsequently bandpass calibrated, the bandpass weights are normalized to unity. But the edges of the coarse channels are noisy and aliased from adjacent coarse channels. Thus, we remove one fine channel from each side of the coarse channel edges from our data. This constitutes $W_f(f)$ in equation~\ref{eqn:obsvis}. The bandpass properties used in this paper are illustrated in figure~\ref{fig:bandpass}.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/bandpass_properties}
\caption{Bandpass properties used in modeling and analysis of delay spectrum of MWA visibilities. The filled circles joined by dotted lines show the 8-tap PFB bandpass shapes obtained using a Kaiser window with parameter $\beta=5$. The {\it plus} symbols joined by dashed lines are the bandpass correction factors. The solid line shows the flagged (gaps) and gain--corrected bandpass shape, $W_f(f)$. These shapes are repeated every 1.28~MHz for the entire bandwidth of 30.72~MHz centered around 185~MHz. \label{fig:bandpass}}
\end{figure}

Thermal noise in simulated visibilities is estimated assuming a system temperature of $T_\textrm{sys}=90$~K per polarization for all frequency channels for all interferometers throughout the course of observation. We take into account the increase in thermal uncertainty in frequency channels that results due to aforementioned bandpass correction.

\section{Observation Parameters}\label{sec:obsparms}

One of the primary targets for EoR observing with the MWA is a patch of sky centered around RA = 00h 00m 00.00s, Dec = -30\arcdeg$\,$00\arcmin$\,$00\farcs 0. This is one of the regions of low foreground emission within 27\arcdeg$\,$ from the zenith. The MWA tracks a patch of sky through antenna beams formed and steered electronically by controlling delay settings of an array of dipoles in a MWA tile. These beamformer delay settings can be changed only in discrete steps. Accordingly, observations were made in a ``drift and shift'' (D-S) mode. Figure~\ref{fig:pointings} shows the direction of pointing of MWA tiles during the course of the observation. The sky is allowed to drift for a certain period of time (usually $\sim 30$~mins) before the discrete shifts in the beamformer delay settings position the beam to be centered again on the same patch of sky. This process is repeated throughout the course of the observation $\approx 4.86$~hours. The LST of two snapshots we chose for our study are shown as vertical lines at -1.92~hours (wrapped by 360\arcdeg from 22.08~hours) and 0.09~hours which are hereafter denoted as {\it off--zenith} and {\it zenith} pointings respectively. These pointings (phased array power patterns) are centered at RA = 4\fdg 387, Dec = -29\fdg 86 and RA = -28\fdg 8, Dec = -26\fdg 701, whereas the visibilities themselves are phased to zenith corresponding to RA = -28\fdg 8, Dec = -26\fdg 701, and RA = 1\fdg 35, Dec = -26\fdg 701 respectively.

@@

Throughout this paper, $\overline{\mathbf{u}}$ and $\overline{\mathbf{x}}$ are assumed to be on a coordinate system aligned with the local east, north (along local meridian) and upward directions at the MWA site. For the conventions of Fourier transform and its inverse we use in this paper, in accordance with equation~\ref{eqn:delay}, signals from the sky towards east and north are recorded with positive delays on eastward and northward oriented antenna spacings respectively. For instance, at the beginning of the observation, the Galactic center is in the westward sky just about to set. Signals from this direction are recorded with a negative delay on interferometers whose antenna spacings are oriented towards the east. Similarly, signals from directions eastward of the local meridian arrive with positive delays on eastward oriented antenna spacings.

For geometrical intuition, we restrict the orientation ($\theta_b$, measured anti-clockwise from East) of all antenna spacings to lie in the range $-67\fdg 5 \leq \theta_b < 112\fdg 5$. Interferometers with antenna spacings oriented in the other half-plane measure conjugate visibilities with delays of equal magnitude but of opposite sign and hence are ignored in our analysis.

\section{Data Analysis}\label{sec:data-analysis}

@@

In order to make a quantitiative comparison of delay spectra obtained with the MWA data and our simulations, we consider the following uncertainties. Our foreground models are derived from other higher frequency catalogs and sky maps. The inherent spread in spectral index increases the uncertainty while predicting fluxes at the observed frequency. Using simple error propagation, the fractional error in the delay spectrum caused by the spread in spectral index is $\sim \ln(f_\textrm{orig}/f)\,\Delta\alpha_\textrm{sp}$, where, $f_\textrm{orig}$ is the original frequency at which the catalog or map was created, $f=185$~MHz is the observed frequency, and $\Delta\alpha_\textrm{sp}$ is the spread (HWHM) in spectral index. From \citet{mau03}, we assume $\Delta\alpha_\textrm{sp} \approx 0.35$ for compact sources from NVSS and SUMSS catalogs. Although the model of \citet{deo08} yields a spectral index per direction on the sky, we assume similar uncertainties exist in spectral indices of our diffuse sky model as well, which is predominantly derived from 408~MHz map of \citet{has82}. Thus, fractional errors in delay spectrum from compact foreground objects and diffuse emission are $\sim$0.7 and $\sim$0.3 respectively. In addition, delay spectra from simulations and data each have uncertainties due to thermal noise with rms, $\Delta V^\textrm{N}_\eta(|\overline{\mathbf{x}}|,\eta) \sim$ 1.4$\times 10^7$~Jy~Hz, where the superscript N stands for thermal noise. We estimate the ratio of delay spectra from data and simulations as, $\rho = |V^\textrm{D}_\eta(|\overline{\mathbf{x}}|,\eta)|\,/\,|V^\textrm{S}_\eta(|\overline{\mathbf{x}}|,\eta)|$, where superscripts D and S denote data and simulation respectively.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/delta_array_histogram_wedge_sim_data_log_ratio_0.3m_ground_custom_gaussian_FG_model_asm_all_sky_nside_64_Tsys_90.0K_185.0_MHz_30.7_MHz_bnw2.0.eps}
\caption{Histogram of logarithm of ratio of observed to simulated visibilities restricted to the {\it foreground wedge} for the {\it off--zenith} (top) and {\it zenith} (bottom) pointings. Median absolute deviations of $\log_{10}\rho$ are 0.29 and 0.27 respectively. A value of zero indicates when the ratio $\rho=1$. \label{fig:data-sim-ratio}}
\end{figure}

@@

\item {\it Interferometer Orientation}: Since the spatial structure of our foreground model is not expected to be isotropic, we divide our interferometer antenna spacings by their orientation ($\theta_\textrm{b}$) measured counter-clockwise from East. We use the following bins: $-67\fdg 5\le \theta_\textrm{b} < -22\fdg 5$, $-22\fdg 5\le \theta_\textrm{b} < 22\fdg 5$, $22\fdg 5\le \theta_\textrm{b} < 67\fdg 5$, and $67\fdg 5\le \theta_\textrm{b} < 112\fdg 5$. The bin centers are oriented towards South-East, East, North-East, and North respectively. Since the delays registered at an interferometer depend on the orientation of its antenna spacing vector, binning by interferometer orientation allows us to attribute the delay spectrum to different sky directions.

@@

It may be noted that numerous features overlap at varying levels of significance as a result of various combinations of parameters. We assign the features to their predominant causes. Secondly, we have used noiseless cases to clearly illustrate the observed foreground signatures. With the addition of noise in the visibilities which is subject to observing time, some of the weaker features may not be as prominently visible. Since the foreground signatures are far too numerous and they reside in a vast volume of parameter space such as interferometer orientation, power pattern, patch of sky under observation, and instrumental configuration, we highlight only some examples of the most notable features observed in the delay spectrum of foreground emission.

@@

Figure~\ref{fig:pitchfork-baselines} shows delay spectra of three interferometers of different antenna spacings oriented northward during the {\it zenith} pointing. The diffuse, compact, and composite components are are shown as dotted, dashed, and solid lines respectively. The horizon delay limits are shown as vertical dot--dashed lines. The gray shaded area denotes the envelope of expected uncertainty in the delay spectrum. This uncertainty is dominated by the spread in spectral index of compact foreground objects inside the {\it foreground} wedge (between the horizon delay limits) proprotional to the delay spectrum from compact objects, and by thermal noise outside the horizon delay limits.

@@

We also note that increasing the antenna spacing results in progressively improving the resolution along delay axis by increasing the number of delay bins inside the {\it foreground wedge}. This improves the localization of foreground objects whose signatures are imprinted in the delay spectrum. For instance, there is an increase in the number of secondary peaks in the delay spectrum between zero--delay and horizon delay limits as antenna spacing increases from $\sim$84~m to $\sim$170~m. In this case, these correspond to sidelobes of the power pattern that lie between the main lobe and the horizon along the North--South direction. In interferometers with short spacings, due to relatively lower resolution along delay axis inside the {\it foreground wedge} and a consequent loss of localization of direction of foreground emission, these secondary peaks blend in with other major peaks and are not distinctly visible. These findings are consistent with dicussion in \S\ref{sec:csm-side-lobe}.

@@

With a foreground model known {\it a priori} in which locations of very bright foreground objects such as the Galactic center or AGN are available, we can predict their response across antenna spacings as a function of observing parameters such as LST, power pattern, etc. Specifically, we can pick interferometers where these bright foreground objects will be visible at delays near horizon delay limits. One such example is already provided in our study. The Galactic center is the most dominant source of foreground contamination in the {\it off--zenith} pointing. Since it is located in the western sky, it is most distinctly seen near the negative horizon delay limit on eastward oriented antenna spacings in the delay spectrum (see figures~\ref{fig:fhd_data}, \ref{fig:sim_data} and \ref{fig:noiseless-dsm-delay-spectrum}, and \S\ref{sec:GC-east}). In fact these figures illustrate that deconvolution along delay axis is severly hampered by this extreme contamination, thereby leading to leakage beyond the {\it foreground wedge} into the {\it EoR window}. In practice, we could downweight these antenna spacings for the specific LST to mitigate the effects of spillover from such foreground contamination in EoR studies. With spatial scales of such objects known, besides downweighting based on antenna spacing orientation, we could also downweight based on antenna spacing distance. For instance, a bright compact object will excite interferometers uniformly across antenna spacings whereas an extended object will gradually fade with increase in antenna spacing.

This provides us with a very simple and yet effective tool in adding a layer of sophistication to mitigate effects of foreground contamination in EoR data analysis. % The success behind clear identification of severely foreground--contaminated interferometers based on delay spectrum technique depends on the localization of the source of foreground contamination and its strength.

% \section{Discussion}\label{sec:discussion}

@@

We also provide a simple and effective tool based on the delay spectrum technique to mitigate foreground contamination in EoR data analysis. It is possible to screen interferometers most affected by foreground contamination as a function of antenna spacing vector and time of observation, which can then be assigned optimal weights in the EoR data analysis chain.

>>>>

Detection of 21~cm H{\sc i} emission from the Epoch of Reionization (EoR), at redshifts $z>6$, is limited primarily by foreground emission. With much of the information coming from the spectral line that maps to distance along line of sight (cosmic time), foregrounds affecting the spectral domain deserve close attention. We investigate in detail the chromatic signatures of an all--sky foreground model imprinted by its spectral structure and instrumental transfer function. We use an antenna pair based delay spectrum technique. Besides confirming foreground contamination is predominantly restricted to a {\it wedge}--shaped region in Fourier space, we demonstrate that the foreground signatures which have the largest impact on the H{\sc i} signal via spillover outside the {\it wedge} arise from power received far away from the primary field of view. Comparing data from recent Murchison Widefield Array (MWA) observations with simulations broken down by sky position and type of emission, we identify diffuse emission near the horizon as the largest contributing factor even on wide antenna spacings relative to compact objects, hitherto unpredicted. These roles are switched in and around the primary field of view. This imprints a {\it pitchfork} on the {\it wedge} in Fourier space. The observation that the brightest contaminants are localized and only effect a fraction of the baselines leads us to propose that by down--weighting certain baselines at certain times, a large fraction of foreground contamination can be mitigated.

@@

Perhaps the key challenge in the statistical detection of power spectrum of the redshifted H{\sc i} signal arises from the contamination by Galactic and extragalactic foregrounds \citep[see, e.g.,][]{dim02,zal04,fur06}. \citet{mor04} show that the inherent isotropy and symmetry of the EoR signal in frequency and spatial wavenumber ($k$) space make it distinguishable from sources of contamination which are isolated to certain $k$ modes by virtue of their inherent spectral smoothness\citep{mor06,bow09,liu11,par12,dil13}. Considerable effort has been made towards understanding the $k$-space behavior foreground signatures in the observed power spectrum \citep{thy13,pob13,mor12,tro12,dat10,bow09}. A general model of how smooth spectrum foregrounds are scattered to higher $k$-modes by the wide-field response of the instrument, into the so-called `wedge', has emerged that provides a high level explanation for the observed foreground power spectrum. The conservative foreground strategy that has developed alongside this work is to discard $k$-modes which could be contaminated (Avoidance). The more aggressive alternative is to subtract a sky model and regain access to modes lost by Avoidance. In both cases, which parts of the sky are most critical to either Avoid or Subtract has remained largely uncertain. Here we will focus primarily on the Avoidance strategy, focusing on identifying foreground components most likely to ``leak'' from foreground modes to EoR modes.

Foregrounds leak into the EoR H{\sc i} power spectrum by either having intrinsic spectral variability, high instrumental chromaticity, or through approximations in power spectrum analysis. 
%Here we focus on the instrumental effects while. 
To decouple from possible analysis effects, such as those pointed out by \citet{haz13}, we compute power spectra using the individual antenna pair based approach of \citet{pob13,par14}.  This approximates the power spectrum as the inverse fourier transform of the spectra generated by the correlator.

 Matching real observations from the Murchison Widefield Array Epoch of Reionization observing program with instrumental simulations of the entire known sky we examine the signature of the power spectrum of each baseline.  We confirm that the foregrounds most likely to obfuscate the Epoch of Reionization power are not caused by emission central field of view but rather from near the horizon. We also find that the smooth galactic emission plays a larger role then previously thought. We quantify this contamination by breaking the simulations down by sky region and type of emission. Using these we arrive at a new method for minimizing the contribution of bright foregrounds which uses prior knowledge of the sky to downweight badly contaminated baselines.

In \S\ref{sec:delay-spectrum} we introduce the delay spectrum technique, then in  \S\ref{sec:instrument}, \S\ref{sec:obsparms}, and \S\ref{sec:data-analysis} we describe the MWA setup, summarize the observing parameters and present the resulting data.  Simulations of these observing parameters are described in \S\ref{sec:modeling} and analyzed by sky location, and type of emission in \S\ref{sec:delay-spectrum-analysis}.  \S\ref{sec:discussion} offers a discussion along with a basic comparison to other current and future instruments. \S\ref{sec:conclusion} offers conclusions and an initial description of a more precise foreground avoidance technique.

@@

In summary, $V_\eta(\overline{\mathbf{u}},\eta)$, is closely related to visibilities, $V_f(\overline{\mathbf{u}},f)$, the basic data blocks measured by each antenna pair, as seen from equation~\ref{eqn:delay-transform}. It captures all the effects of EoR H{\sc i} signal corruption caused by foregrounds and the instrument. At the same time, it is very closely related to the sought power $|V(\overline{\mathbf{k}}_\perp,k_\parallel)|^2$ in $k$--modes containing critical information about spatial scales. For our analysis, we use an equivalent quantity, $V_\eta(\overline{\mathbf{x}},\eta) \equiv V_\eta(\lambda\overline{\mathbf{u}},\eta)$. For purposes of illustration, we often use a simplified quantity $|V_\eta(|\overline{\mathbf{x}}|,\eta)|$.

@@

\section{Murchison Widefield Array Observations}\label{sec:instrument}

\subsection{The MWA}

The MWA was completed in 2012, and after commissioning, began its EoR observing program in 2013. In its final configuration the MWA is a 128--tile interferometer array capable of observing a 30~MHz band anywhere in the range 80--300~MHz.  Each tile is a phased array of 16 dipoles each in the shape of a bow--tie characterized by a grating response with a $\sim$~25\arcdeg wide primary field of view and multiple secondary side lobes. The tile arrangement is a centrally condensed core of $\sim$~300~m --- there are many spacings in the range 5--50~m --- and a radial density that falls off as the inverse of the radius with the longest baselines at 3~km \citep{bea12}. Here we focus on antenna spacings $|\overline{\mathbf{x}}| \le 200$~m (spatial scales relevant to reionization) and a frequency range 170--200~MHz.

One important feature of the MWA's measured spectrum is the shape of its bandpass. The band is divided coarsely into 24 chunks with each chunk weighted by a digital filter. The channels at the edges of these chunks have been flagged because they are known to possibly be contaminated by aliasing at a low level. Any spectral flags, or other known weights constitutes $W_f(f)$ in equation~\ref{eqn:obsvis}. The bandpass properties used in this paper are illustrated in figure~\ref{fig:bandpass}.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/bandpass_properties}
\caption{Bandpass properties used in modeling and analysis of delay spectrum of MWA visibilities. The filled circles joined by dotted lines show the 8--tap PFB bandpass shapes obtained using a Kaiser window with parameter $\beta=5$. The {\it plus} symbols joined by dashed lines are the bandpass correction factors. The solid line shows the flagged (gaps) and gain--corrected bandpass shape, $W_f(f)$. These shapes are repeated every 1.28~MHz for the entire bandwidth of 30.72~MHz centered around 185~MHz. \label{fig:bandpass}}
\end{figure}

Thermal noise in simulated visibilities is estimated assuming a system temperature of $T_\textrm{sys}=90$~K per polarization for all frequency channels for all antenna pairs throughout the course of observation. We take into account the increase in thermal uncertainty in frequency channels that results due to aforementioned bandpass correction. ({\bf Since the bandpass correction is digital, shouldn't the noise stay flat after bandpass correction?})

\subsection{Observation Parameters}\label{sec:obsparms}

One of the primary targets for EoR observing with the MWA is a patch of sky centered around RA = 00h 00m 00.00s, Dec = -30\arcdeg$\,$00\arcmin$\,$00\farcs 0. This is one of the regions of low foreground emission within 27\arcdeg$\,$ from the zenith. The MWA tracks a patch of sky through antenna beams formed and steered electronically by controlling delay settings of an array of dipoles in a MWA tile. These beamformer delay settings can be changed only in discrete steps. The sky is allowed to drift for a certain period of time (usually $\sim 30$~mins) before the discrete shifts in the beamformer delay settings position the beam to be centered again on the same patch of sky. This process is repeated throughout the course of the observation $\approx 4.86$~hours. Figure~\ref{fig:pointings} illustrates how the apparent pointing oscillates around the desired pointing direction.

The LST of two snapshots we chose for our study are shown as vertical lines at -1.92~hours (i.e., 22.08~hours) and 0.09~hours which are hereafter denoted as {\it off--zenith} and {\it zenith} pointings respectively. These pointings (phased array power patterns) are centered at RA = 4\fdg 387, Dec = -29\fdg 86 and RA = -28\fdg 8, Dec = -26\fdg 701, whereas the visibilities themselves are phased to zenith corresponding to RA = -28\fdg 8, Dec = -26\fdg 701, and RA = 1\fdg 35, Dec = -26\fdg 701 respectively.

The observations reported here were observed, near the beginning of the campaign, on Aug 23, 2013. This day has been chosen as a reference data set for comparison within the MWA EoR collaboration. This day was selected on the basis of low number of known system errors from among a long stretch of stable observing.

@@

\subsection*{A note on baselines/delay spectra sign convention}

Throughout this paper, $\overline{\mathbf{u}}$ and $\overline{\mathbf{x}}$ are assumed to be on a coordinate system aligned with the local east, north (along local meridian) and upward directions at the MWA site. For the conventions of Fourier transform and its inverse we use in this paper, in accordance with equation~\ref{eqn:delay}, signals from the sky towards east and north are recorded with positive delays on eastward and northward oriented antenna spacings respectively. For instance, at the beginning of the observation, the Galactic center is in the westward sky just about to set. Signals from this direction are recorded with a negative delay on baselines whose antenna spacings are oriented towards the east. Similarly, signals from directions eastward of the local meridian arrive with positive delays on eastward oriented antenna spacings.

For geometrical intuition, we restrict the orientation ($\theta_b$, measured anti-clockwise from East) of all antenna spacings to lie in the range $-67\fdg 5 \leq \theta_b < 112\fdg 5$. Antenna pairs with spacings oriented in the other half--plane measure conjugate visibilities with delays of equal magnitude but of opposite sign and hence are ignored in our analysis.

\subsection{Data Analysis}\label{sec:data-analysis}

The data are flagged for interference \citep{off12,off10}, removing $\sim 3$\% of the data. These are averaged in time and frequency to 2s and 40kHz respectively.

@@

In order to make a quantitiative comparison of delay spectra obtained with the MWA data and our simulations, we consider the following uncertainties. Our foreground models are derived from other higher frequency catalogs and sky maps. The inherent spread in spectral index increases the uncertainty while predicting fluxes at the observed frequency. Using simple error propagation, the fractional error in the delay spectrum caused by the spread in spectral index is $\sim \ln(f_\textrm{orig}/f)\,\Delta\alpha_\textrm{sp}$, where, $f_\textrm{orig}$ is the original frequency at which the catalog or map was created, $f=185$~MHz is the observed frequency, and $\Delta\alpha_\textrm{sp}$ is the spread (HWHM) in spectral index. From \citet{mau03}, we assume $\Delta\alpha_\textrm{sp} \approx 0.35$ for compact sources from NVSS and SUMSS catalogs. Although the model of \citet{deo08} yields a spectral index per direction on the sky, we assume similar uncertainties exist in spectral indices of our diffuse sky model as well, which is predominantly derived from 408~MHz map of \citet{has82}. Thus, fractional errors in delay spectrum from compact foreground objects and diffuse emission are $\sim$0.7 and $\sim$0.3 respectively. In addition, delay spectra from simulations and data each have uncertainties due to thermal noise with rms, $\Delta V^\textrm{N}_\eta(|\overline{\mathbf{x}}|,\eta) \sim$ 1.4$\times 10^7$~Jy~Hz, where the superscript N stands for thermal noise. We estimate the ratio of delay spectra from data and simulations as, $\rho = |V^\textrm{D}_\eta(|\overline{\mathbf{x}}|,\eta)|\,/\,|V^\textrm{S}_\eta(|\overline{\mathbf{x}}|,\eta)|$, where superscripts D and S denote data and simulation respectively. % Using the aforementioned uncertainties added in quadrature, we estimate the approximate expected uncertainty in $\log_{10}\rho$ and denote it by $\Delta\,\log_{10}\rho$.

% Figure~\ref{fig:data-sim-ratio} shows the significance of $\log_{10}\rho$ relative to the expected error $\Delta\,\log_{10}\rho$ for all $|\overline{\mathbf{x}}|$ and $\eta$ inside the {\it foreground wedge}. The probability densities and cumulative distributions are shown in black and gray respectively. Also shown are dotted vertical lines which denote equality between $\Delta\,\log_{10}\rho$ and $\log_{10}\rho$. The fraction of the {\it foreground wedge} that lies in between these vertical lines is $\sim$~90\% and $\sim$~95\% for {\it off--zenith} and {\it zenith} pointings respectively. In other words, subject to expected uncertainties, the ratio of delay spectra obtained with MWA data and simulations are consistent with unity at $\sim$~90\% and $\sim$~95\% confidence levels in the {\it off--zenith} and {\it zenith} pointings respectively.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/delta_array_histogram_wedge_sim_data_log_ratio_0.3m_ground_custom_gaussian_FG_model_asm_all_sky_nside_64_Tsys_90.0K_185.0_MHz_30.7_MHz_bnw2.0.eps}
\caption{Histogram of logarithm of ratio of observed to simulated visibilities restricted to the {\it foreground wedge} for the {\it off--zenith} (top) and {\it zenith} (bottom) pointings. Median absolute deviations of $\log_{10}\rho$ are 0.29 and 0.27 respectively. A value of zero indicates when the ratio $\rho=1$. \label{fig:data-sim-ratio}}
% \includegraphics[width=\linewidth]{figures/v1_0/delta_array_histogram_wedge_sim_data_snr_log_ratio_0.3m_ground_custom_gaussian_FG_model_asm_all_sky_nside_64_Tsys_90.0K_185.0_MHz_30.7_MHz_bnw2.0.eps}
% \caption{Significance of logarithm of ratio of observed to simulated visibilities relative to expected uncertainty restricted to the {\it foreground wedge} for the {\it off--zenith} (top) and {\it zenith} (bottom) pointings. Probability densities and cumulative distributions are shown in black and gray respectively. Region inside the dotted vertical lines denotes when the ratio, $\rho$, is consistent with unity subject to expected uncertainty. Also indicated in percentages are confidence levels estimated conservatively for such a consistency. They are $\sim$~90\% and $\sim$~95\% for {\it off--zenith} and {\it zenith} pointings respectively. \label{fig:data-sim-ratio-snr}}
\end{figure}

@@

\item {\it Antenna Pair Orientation}: Since the spatial structure of our foreground model is not expected to be isotropic, we divide our antenna pairs by their orientation ($\theta_\textrm{b}$) measured counter-clockwise from East. We use the following bins: $-67\fdg 5\le \theta_\textrm{b} < -22\fdg 5$, $-22\fdg 5\le \theta_\textrm{b} < 22\fdg 5$, $22\fdg 5\le \theta_\textrm{b} < 67\fdg 5$, and $67\fdg 5\le \theta_\textrm{b} < 112\fdg 5$. The bin centers are oriented towards South-East, East, North-East, and North respectively. Since delays depend on antenna pair orientation, binning by interferometer orientation allows us to match delay spectrum features to different sky directions.

@@

It may be noted that numerous features overlap at varying levels of significance as a result of various combinations of parameters. We assign the features to their predominant causes. Secondly, we have used noiseless cases to clearly illustrate the observed foreground signatures. With the addition of noise in the visibilities which is subject to observing time, some of the weaker features may not be as prominently visible. Since the foreground signatures are far too numerous and they reside in a vast volume of parameter space such as antenna pair orientation, power pattern, patch of sky under observation, and instrumental configuration, we highlight only some examples of the most notable features observed in the delay spectrum of foreground emission.

@@

Figure~\ref{fig:pitchfork-baselines} shows delay spectra of three antenna pairs of different antenna spacings oriented northward during the {\it zenith} pointing. The diffuse, compact, and composite components are are shown as dotted, dashed, and solid lines respectively. The horizon delay limits are shown as vertical dot--dashed lines. The gray shaded area denotes the envelope of expected uncertainty in the delay spectrum. This uncertainty is dominated by the spread in spectral index of compact foreground objects inside the {\it foreground} wedge (between the horizon delay limits) proprotional to the delay spectrum from compact objects, and by thermal noise outside the horizon delay limits.

@@

We also note that increasing the antenna spacing results in progressively improving the resolution along delay axis by increasing the number of delay bins inside the {\it foreground wedge}. This improves the localization of foreground objects whose signatures are imprinted in the delay spectrum. For instance, there is an increase in the number of secondary peaks in the delay spectrum between zero--delay and horizon delay limits as antenna spacing increases from $\sim$84~m to $\sim$170~m. In this case, these correspond to sidelobes of the power pattern that lie between the main lobe and the horizon along the North--South direction. In short antenna spacings, due to relatively lower resolution along delay axis inside the {\it foreground wedge} and a consequent loss of localization of direction of foreground emission, these secondary peaks blend in with other major peaks and are not distinctly visible. These findings are consistent with dicussion in \S\ref{sec:csm-side-lobe}.

@@

With a foreground model known {\it a priori} in which locations of very bright foreground objects such as the Galactic center or AGN are available, we can predict their response across antenna spacings as a function of observing parameters such as LST, power pattern, etc. Specifically, we can pick antenna pairs where these bright foreground objects will be visible at delays near horizon delay limits. One such example is already provided in our study. The Galactic center is the most dominant source of foreground contamination in the {\it off--zenith} pointing. Since it is located in the western sky, it is most distinctly seen near the negative horizon delay limit on eastward oriented antenna spacings in the delay spectrum (see figures~\ref{fig:fhd_data}, \ref{fig:sim_data} and \ref{fig:noiseless-dsm-delay-spectrum}, and \S\ref{sec:GC-east}). In fact these figures illustrate that deconvolution along delay axis is severly hampered by this extreme contamination, thereby leading to leakage beyond the {\it foreground wedge} into the {\it EoR window}. In practice, we could downweight these antenna spacings for the specific LST to mitigate the effects of spillover from such foreground contamination in EoR studies. With spatial scales of such objects known, besides downweighting based on antenna spacing orientation, we could also downweight based on antenna spacing distance. For instance, a bright compact object will yield uniform {\it visibility} amplitudes across antenna spacings whereas that of an extended object will gradually fade with increase in antenna spacing.

This provides us with a very simple and yet effective tool in adding a layer of sophistication to mitigate effects of foreground contamination in EoR data analysis. % The success behind clear identification of severely foreground--contaminated antenna pairs based on delay spectrum technique depends on the localization of the source of foreground contamination and its strength.

\section{Discussion}\label{sec:discussion}

\section{Conclusion}\label{sec:conclusion}

@@

We also provide a simple and effective tool based on the delay spectrum technique to mitigate foreground contamination in EoR data analysis. It is possible to screen antenna pairs most affected by foreground contamination as a function of antenna spacing vector and time of observation, which can then be assigned optimal weights in the EoR data analysis chain.
